The Role of Ice Compositions and Morphology for Snowlines and the C/N/O Ratios in Active Disks

### 2016-01-25 tex/draft2/ms.tex
update

Following \citet{chiang10}, the temperature profile for a static disk is
\begin{equation}
\label{eq:diskT}
T = 120\, (r/\text{AU})^{-3/7} \,\,\text{K},
\end{equation}
%We model the static disk as a minimum mass solar nebula (MMSN), using a prescription for the gas surface density, $\Sigma$, and disk midplane temperature, $T$, similar to that of \citet{chiang10}:
%\begin{subeqnarray}
%\label{eq:disk}
%\Sigma&=&2000\, (r/\text{AU})^{-1}\,\, \text{g cm}^{-2} \slabel{eq:disksigma}\\
%T &=& 120\, (r/\text{AU})^{-3/7} \,\,\text{K}, \slabel{eq:diskT}
%\end{subeqnarray}
where $r$ is the semimajor axis.% Based on observations of protoplanetary disks \citep{andrews10}, we choose a flatter surface density than the one assumed by \citet{chiang10}, where $\Sigma \propto r^{-3/2}$. 
We use the \citet{shakura73}  steady-state disk solution to model the viscous disk. From Paper I, the viscous disk temperature profile is computed as % Solving the Equation set of Appendix A in Paper I yields an expression for the temperature profile in a steady-state disk:
\begin{equation}
\label{eq:Tdact}
T_{\rm act}=\frac{1}{4 r} \Big(\frac{3 G \kappa_0\dot{M}^2 M_* \mu m_{\rm p} \Omega_{\rm k}}{\pi^2 \alpha k_{\rm B} \sigma}\Big)^{1/3}.
\end{equation}
Here $G$ is the gravitational constant, $\kappa_0=2 \times 10^{-6}$ is a dimensionless coefficient in the opacity law $\kappa=\kappa_0 T_{\rm act}^2$, $M_*=M_{\odot}$ is the mass of the central star, $\mu=2.35$ is the mean molecular weight of the nebular gas, $m_{\rm p}$ is the proton mass, $\Omega_{\rm k}=\sqrt{G M_{\odot}/r^3}$ is the Keplerian angular velocity, $\alpha=0.01$ is a dimensionless coefficient (see below for details), $k_{\rm B}$ is the Boltzmann constant, and $\sigma$ is the Stefan-Boltzmann constant. The final midplane temperature profile is computed as 
\begin{equation}
\label{eq:activeT}
T^4 = T_{\rm act}^4 + T_{\rm irr}^4,
\end{equation}
where $T_{\rm irr}=T$ from Equation (\ref{eq:diskT}). We use this expression because in addition to accretion heating, stellar irradiation also contributes to the disk thermal structure.

The steady-state disk has an $\alpha$-viscosity prescription, where the kinematic viscosity is $\mu=\alpha c H$. Here $c \equiv \sqrt{k_{\rm B} T /(\mu m_{\rm p})}$ is the isothermal sound speed (with $T$ from Equation \ref{eq:activeT}), and $H \equiv c/\Omega_{\rm k}$ is the disk scale height. We can then determine the gas surface density for a viscous disk as (\citealt{shakura73}; see also Paper I for a more detailed explanation of these calculations):
\begin{equation}
\label{eq:Sigmaact}
\Sigma=\frac{\dot{M}}{3 \pi \nu}.
\end{equation}
We choose $\dot{M}=10^{-8} M_{\odot}$ yr$^{-1}$, consistent with mass flux observations in disks (e.g., \citealt{andrews10}). As acknowledged in Paper I, the mass flux rate $\dot{M}$ and stellar luminosity $L_*$ will vary throughout the disk lifetime (\citealt{kennedy06}, \citealt{chambers09}), in contrast with our simplified model which assumes that both quantities are constant. This effect will be most pronounced in the inner disk ($\lesssim$ few AU), where accretion heating dominates. We thus acknowledge that the location of the H$_2$O snowline may be determined by the decline in $\dot{M}$ or $L_*$ with time, rather than radial drift (see Paper I, Section 2.1 for a more detailed explanation).

@@

\section{Discussion}
\label{sec:discussion}

\emgr{Discuss how entrapment of volatiles by H2O affects volatile abundances and C/O ratios. Re-emphasize the fact that the C/O and N/O ratios are upper estimates, and that CH4 and NH3 might matter in a viscous disk. State that we plan to address this in a future paper. More TBD.}

@@

\emgr{Maybe we can include the summary in the discussion section?}

>>>>

Following \citet{chiang10}, the temperature profile for a static disk is
\begin{equation}
\label{eq:diskT}
T = 120\, (r/\text{AU})^{-3/7} \,\,\text{K},
\end{equation}
%We model the static disk as a minimum mass solar nebula (MMSN), using a prescription for the gas surface density, $\Sigma$, and disk midplane temperature, $T$, similar to that of \citet{chiang10}:
%\begin{subeqnarray}
%\label{eq:disk}
%\Sigma&=&2000\, (r/\text{AU})^{-1}\,\, \text{g cm}^{-2} \slabel{eq:disksigma}\\
%T &=& 120\, (r/\text{AU})^{-3/7} \,\,\text{K}, \slabel{eq:diskT}
%\end{subeqnarray}
where $r$ is the semimajor axis.% Based on observations of protoplanetary disks \citep{andrews10}, we choose a flatter surface density than the one assumed by \citet{chiang10}, where $\Sigma \propto r^{-3/2}$. 
We use the \citet{shakura73}  steady-state disk solution to model the viscous disk. From Paper I, the viscous disk temperature profile is computed as % Solving the Equation set of Appendix A in Paper I yields an expression for the temperature profile in a steady-state disk:
\begin{equation}
\label{eq:activeT}
T^4 =\Big[\frac{1}{4 r} \Big(\frac{3 G \kappa_0\dot{M}^2 M_* \mu m_{\rm p} \Omega_{\rm k}}{\pi^2 \alpha k_{\rm B} \sigma}\Big)^{1/3}\Big]^4 + T_{\rm irr}^4,
\end{equation}
where $T_{\rm irr}=T$ from Equation (\ref{eq:diskT}). Here $G$ is the gravitational constant, $\kappa_0=2 \times 10^{-6}$ is a dimensionless opacity coefficient, $M_*=M_{\odot}$ is the mass of the central star, $\mu=2.35$ is the mean molecular weight of the nebular gas, $m_{\rm p}$ is the proton mass, $\Omega_{\rm k}=\sqrt{G M_{\odot}/r^3}$ is the Keplerian angular velocity, $\alpha=0.01$ is a dimensionless coefficient (see below for details), $k_{\rm B}$ is the Boltzmann constant, and $\sigma$ is the Stefan-Boltzmann constant.

%The final midplane temperature profile is computed as 
%\begin{equation}
%\label{eq:activeT}
%T^4 = T_{\rm act}^4 + T_{\rm irr}^4,
%\end{equation}
%where $T_{\rm irr}=T$ from Equation (\ref{eq:diskT}). We use this expression because in addition to accretion heating, stellar irradiation also contributes to the disk thermal structure.

The steady-state disk has an $\alpha$-viscosity prescription, where the kinematic viscosity is $\nu=\alpha c H$. Here $c \equiv \sqrt{k_{\rm B} T /(\mu m_{\rm p})}$ is the isothermal sound speed (with $T$ from Equation \ref{eq:activeT}), and $H \equiv c/\Omega_{\rm k}$ is the disk scale height. We can then determine the gas surface density for a viscous disk as (\citealt{shakura73}; see also Paper I for a more detailed explanation of these calculations):
\begin{equation}
\label{eq:Sigmaact}
\Sigma=\frac{\dot{M}}{3 \pi \nu}.
\end{equation}
We choose $\dot{M}=10^{-8} M_{\odot}$ yr$^{-1}$, consistent with mass flux observations in disks (e.g., \citealt{andrews10}). As acknowledged in Paper I, the mass flux rate $\dot{M}$ and stellar luminosity $L_*$ will vary throughout the disk lifetime (\citealt{kennedy06}, \citealt{chambers09}), in contrast with our simplified model which assumes that both quantities are constant. This effect will be most pronounced in the inner disk ($\lesssim$ few AU), where accretion heating dominates. We thus acknowledge that the location of the H$_2$O snowline may be determined by the decline in $\dot{M}$ or $L_*$ with time, rather than radial drift (see Paper I, Section 2.1 for a more detailed explanation).

@@

%\section{Discussion}
%\label{sec:discussion}

%\emgr{Discuss how entrapment of volatiles by H2O affects volatile abundances and C/O ratios. Re-emphasize the fact that the C/O and N/O ratios are upper estimates, and that CH4 and NH3 might matter in a viscous disk. State that we plan to address this in a future paper. More TBD.}

@@

In this paper we explore the role of icy grain morphology and disk dynamics on the snowline locations of major volatile carrier molecules and the C/N/O ratios in protoplanetary disks. We enhance the coupled drift-desorption model developed in \citet{piso15b} by adding more carbon- and nitrogen-bearing species into our framework and by considering different environments in which the icy grains reside. Our results can be summarized as follows:

\begin{enumerate}

\item Due to the high volatility of N$_2$, the gaseous N/O ratio in the outer disk is enhanced by more than a factor of three compared to its average value. This enhancement is more pronounced than in the case of the gas-phase C/O ratio, which is increased by a factor of two compared to the stellar value. Moreover, the N/O ratio in gas is expected to be very large between the CO and N$_2$ snowlines due to the complete depletion of oxygen gas in this region

\item The presence of some carbon in the form of CH$_4$ and of some nitrogen in the form of NH$_3$ only modestly affects our results for the C/O and N/O ratios, respectively. In both cases, large C/O and N/O ratios in the outer disk are preserved.

\item Grain composition sensitively affects the CO and N$_2$ snowline locations. If CO and N$_2$ are layered on a water-ice substrate rather than existing as pure ices, their snowlines move inward by up to $\sim$70 \%. This effect is separate from that of radial drift and viscous gas accretion, which also cause an inward movement of the snowlines by up to $\sim$50 \%.

\end{enumerate}

Our results have direct consequences for the composition of nascent giant planets. The considerable inward movement of the CO and N$_2$ snowlines due to the ice grains residing in a water ice environment rather than as pure ices implies than giant planets with high C/O and/or N/O ratios in their atmospheres may form closer in than previously predicted by theoretical models. Moreover, our model shows that wide separation gas giants may have an excess of nitrogen in their envelopes, which may be used to trace their origins. In future work, we plan to add new levels of complexity to our model in terms of disk chemistry, dynamics, and planetary dynamics, thus forming a solid framework for understanding the origins of gas giants.

%\emgr{Maybe we can include the summary in the discussion section?}
